Topology of the $O(3)$ non-linear sigma model under the gradient flow

### 2021-11-10 proceedings/draft.tex
Incorporated Daniel's paper and removed general comments

% Please make sure you insert your
% data according to the instructions in PoSauthmanual.pdf
\documentclass[a4paper,11pt]{article}
\usepackage{pos}
% \usepackage{eulervm}
% \usepackage{xspace}
% \usepackage{inputenc}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{subcaption}
\newcommand{\e}{\vec e}
\newcommand{\logchi}{7.4 }
\newcommand{\powchi}{8.8 }
\graphicspath{{../figures/}}
\newcommand\chris[1]{\textcolor{teal}{CJM: #1}}
\newcommand\stuart[1]{\textcolor{pink}{SNT: #1}}
% \newcommand\chris[1]{}
% \newcommand\stuart[1]{}

@@

% \affiliation[b]{Thomas Jefferson National Accelerator Facility, Newport News, Virginia, USA}

@@

\abstract{
The $O(3)$ non-linear sigma model (NLSM) is a prototypical field theory for QCD and ferromagnetism, featuring topological qualities. Though the topological susceptibility $\chi_t$ should vanish in physical theories, lattice simulations of the NLSM find that $\chi_t$ diverges in the continuum limit. We study the effect of the gradient flow on this quantity using a Markov Chain Monte Carlo method, finding that a logarithmic divergence persists. This result supports a previous study and indicates that either the definition of topological charge is problematic or the NLSM has no well-defined continuum limit. We also introduce a $\theta$-term and analyze the topological charge as a function of $\theta$ under the gradient flow. \chris{May need to adjust this to new reference.}}

@@

\chris{TO DO:}
\begin{enumerate}
 \item \chris{Stuart: Address comments in figures.}
 \item \chris{Chris: Incorporate new reference from conference.}
 \item \chris{Chris: Double check fits with e.g. lsqfit}
\end{enumerate}

@@

The protection of topological instantons in the 1+1 $O(3)$ NLSM relies on a vanishing topological susceptibilty. However, the convergence of this quantity is still unknown \cite{bietenholz2018}. Analytical arguments suggest the topological susceptibility should approach zero in the continuum limit, but numerical results on the lattice predict infinities \cite{berg1981}. To shed light on this apparent contradiction, we apply the gradient flow, a local smearing of operators which preserves gauge invariance. In quantum chromodynamics, this technique has corroborated a previous analytical result \cite{giusti2004} by removing ultraviolet divergences on the lattice \cite{bruno2014}. This success has motivated the gradient flow to calculate the topological susceptibility in the 1+1 $O(3)$ NLSM. Despite this intuition, recent studies demonstrate that the observable still diverges in the continuum limit \cite{bietenholz2018}.

We also study a second perspective on the topological susceptibility arising from the introduction of a $\theta$-term into the field Lagrangian. This term drives the vacuum state into a topological phase \cite{allessalom2008}. Differentiating the field's partition function with respect to $\theta$ yields a value proportional to the topological susceptibility. The effect of nonzero $\theta$ on the theory therefore should reflect the divergence in the continuum limit. In this work we verify the divergence of the topological susceptibility and develop a clearer picture of how the $\theta$-term affects the topology of the 1+1 $O(3)$ NLSM.

@@

From the definition of $Q$, we define the topological susceptibility as
\begin{equation}
\chi_t \equiv \frac{1}{L^2} \Big( \langle Q^2 \rangle - \langle Q \rangle^2 \Big).
\end{equation}
Since $\langle Q \rangle=0$ in the NLSM, this quantity becomes
\begin{equation}
    \chi_t = \frac{1}{L^2} \sum_{x^*} \langle q(x^*)q(0)\rangle,
\end{equation}
where we have assumed periodic boundary conditions. In the continuum limit, the topological susceptibility diverges owing solely to the $x^*=0$ term \cite{bietenholz2018}, a divergence that also exists in QCD \cite{bruno2014}.

@@

We thermalize the configurations with 1000 sweeps, with a cluster update every five sweeps, and illustrate a sample Markov chain in Fig.~\ref{fig:therm}, where we plot the action as a function of Metropolis sweeps. We use Wolff's automatic windowing procedure \cite{wolff2007} to estimate the autocorrelation times for various observables, such as the magnetic susceptibility $\chi_m$. We measure observables every $50$ sweeps for each simulation.
\begin{figure}[h]
  \centering
      \begin{subfigure}[b]{0.5\textwidth}\centering
        \includegraphics[width=\textwidth]{therm24.png}
        \caption{$L=24$}
      \end{subfigure}%
      \hfill
      \begin{subfigure}[b]{0.5\textwidth}\centering
        \includegraphics[width=\textwidth]{therm404.png}
        \caption{$L=404$}
      \end{subfigure}
      \hfill
      \caption{\label{fig:therm} Plots of the action as a function of Monte Carlo time, starting with a random NLSM lattice.}
\end{figure}

@@

We check our numerical implementation of the NLSM by comparison to studies of the internal energy and magnetic susceptibility in Refs.~\cite{berg1981} and \cite{bietenholz2018}. Following \cite{berg1981}, we approximate the internal energy in the strong ($\beta<1$) and weak ($\beta>2$) regimes as
\begin{equation}
    \label{eq:analytic_energy}
E \approx \begin{cases}
    4-4y-8y^3-\frac{48}{5}y^5& \beta<1 \\
\frac{2}{\beta} + \frac{4}{\beta^2} + 0.156\frac{1}{\beta^3}& \beta>2,
\end{cases}
\end{equation}
where
\begin{equation}
    y=\mathrm{coth}\beta-\frac{1}{\beta}.
\end{equation}
We compare this analytical result and simulated values of $\chi_m$ with the Monte Carlo simulation in Fig.~\ref{fig:bergluscher}.
\begin{figure}[h!]
    \centering
      % \includegraphics[width=0.45\textwidth]{int_energy-0.png}
      % \includegraphics[width=0.45\textwidth]{int_energy-1.png}
      \includegraphics[width=0.9\textwidth]{berg_luscher.png}
      \caption{\label{fig:bergluscher} Comparison with \cite{berg1981}. First panel: internal energy compared with analytic  energy (Eq.~\ref{eq:analytic_energy}). Second panel: magnetic susceptibility compared with literature values.}
\end{figure}
The slight discrepancy between numerical results in the cross-over region provides an estimate of unquantified systematic uncertainties.

@@

\section{Discussion}

\chris{This discussion will need to be updated to incorporate the new reference from our lattice conference discussions.}

Berg \& L\"uscher \cite{berg1981} originally illuminated this discrepancy between the renormalization group hypothesis (that $\chi_t \rightarrow 0$) and the numerical results, providing three possible causes:
\begin{enumerate}
    \item The definition of the topological charge does not scale to the continuum.
    \item There are ultraviolet divergences.
    \item There is no reasonable continuum limit.
\end{enumerate}
Since the gradient flow suppresses ultraviolet fluctuations, the persistence of a divergent topological susceptibility under the gradient flow undermines the second option, thereby supporting the other two.

% Further discussions\footnote{We would like to thank Daniel Nogradi for these conference discussions} \cite{nogradi2012}

>>>>

% Please make sure you insert your
% data according to the instructions in PoSauthmanual.pdf
\documentclass[a4paper,11pt]{article}
\usepackage{pos}
% \usepackage{eulervm}
% \usepackage{xspace}
% \usepackage{inputenc}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{subcaption}
\usepackage{enumitem}
\newcommand{\e}{\vec e}
\newcommand{\logchi}{7.4 }
\newcommand{\powchi}{8.8 }
\graphicspath{{../figures/}}
\newcommand\chris[1]{\textcolor{teal}{CJM: #1}}
\newcommand\stuart[1]{\textcolor{pink}{SNT: #1}}
% \newcommand\chris[1]{}
% \newcommand\stuart[1]{}

@@

@@

\abstract{
The $O(3)$ non-linear sigma model (NLSM) is a prototypical field theory for QCD and ferromagnetism, featuring topological qualities. Though the topological susceptibility should vanish in physical theories, lattice simulations of the NLSM find that the topological susceptibility diverges in the continuum limit. We introduce a $\theta$-term and analyze the topological charge as a function of $\theta$ under the gradient flow. Our results show that divergence persists in the presence of the flow, even at non-zero $\theta$.
}

@@

\chris{TO DO:}
\begin{enumerate}
 \item \chris{Stuart: check reference highlighted below in introduction}
 \item \chris{Chris: Double check fits with e.g. lsqfit}
\end{enumerate}

@@

The protection of topological instantons in the 1+1 $O(3)$ NLSM relies on a vanishing topological susceptibilty. However, the convergence properties of this quantity are still unclear \cite{bietenholz2018}. Analytical arguments suggest the topological susceptibility should approach zero in the continuum limit \cite{berg1981} \chris{Can you please confirm this is the correct reference for the origin of these ``analytical arguments''?}, but numerical results on the lattice, summarized in \cite{bietenholz2018}, indicate a logarithmic divergence and support the semi-classical picture of small-size instantons generating the divergence. To elucidate this apparent contradiction, we apply the gradient flow, a local smearing of operators which preserves gauge invariance. In quantum chromodynamics, this technique has corroborated a previous analytical result \cite{giusti2004} by removing ultraviolet divergences on the lattice \cite{bruno2014}. This success has motivated the gradient flow to as a tool to define the continuum limit of the topological susceptibility in the 1+1 $O(3)$ NLSM. Despite this intuition, recent studies demonstrate that the topological susceptibility still diverges in the continuum limit \cite{bietenholz2018}.

We also study a second perspective on the topological susceptibility arising from the introduction of a $\theta$-term into the field Lagrangian. This term drives the vacuum state into a topological phase \cite{allessalom2008}, and the resulting theory exhibits confining, walking, and conformal behaviors at different values of $\theta$ \cite{Nogradi:2012dj}. Differentiating the partition function with respect to $\theta$ yields a value proportional to the topological susceptibility. The effect of nonzero $\theta$ on the theory therefore should reflect the divergence of the susceptibility in the continuum limit. In this work we verify the divergence of the topological susceptibility and develop a clearer picture of how the $\theta$-term affects the topology of the 1+1 $O(3)$ NLSM.

@@

From the definition of $Q$, we define the topological susceptibility as
\begin{equation}
\chi_t \equiv \frac{1}{L^2} \Big( \langle Q^2 \rangle - \langle Q \rangle^2 \Big).
\end{equation}
Since $\langle Q \rangle=0$ in the NLSM, this quantity becomes
\begin{equation}
    \chi_t = \frac{1}{L^2} \sum_{x^*} \langle q(x^*)q(0)\rangle,
\end{equation}
where we have assumed periodic boundary conditions. In the continuum limit, the topological susceptibility diverges owing solely to the $x^*=0$ term \cite{bietenholz2018}, a divergence that also exists in QCD \cite{bruno2014}. In the semiclassical approximation, this divergence arises from small-size instantons \cite{Luscher:1981tq}.

@@

We thermalize the configurations with 1000 sweeps, with a cluster update every five sweeps, and illustrate a sample Markov chain in Fig.~\ref{fig:therm}, where we plot the action as a function of Metropolis sweeps. We use Wolff's automatic windowing procedure \cite{wolff2007} to estimate the autocorrelation times for various observables, such as the magnetic susceptibility $\chi_m$. We measure observables every $50$ sweeps for each simulation.
\begin{figure}[h]
  \centering
      \begin{subfigure}[b]{0.5\textwidth}\centering
        \includegraphics[width=\textwidth]{therm24.png}
        \caption{$L=24$}
      \end{subfigure}%
      \hfill
      \begin{subfigure}[b]{0.5\textwidth}\centering
        \includegraphics[width=\textwidth]{therm404.png}
        \caption{$L=404$}
      \end{subfigure}
      \hfill
      \caption{\label{fig:therm} Plots of the action, $S$, as a function of Monte Carlo time, starting with a random NLSM lattice.}
\end{figure}

@@

We check our numerical implementation of the NLSM by comparison to studies of the internal energy and magnetic susceptibility in Refs.~\cite{berg1981} and \cite{bietenholz2018}. Following \cite{berg1981}, we approximate the internal energy in the strong ($\beta<1$) and weak ($\beta>2$) regimes as
\begin{equation}
    \label{eq:analytic_energy}
E \approx \begin{cases}
    4-4y-8y^3-\frac{48}{5}y^5& \beta<1 \\
\frac{2}{\beta} + \frac{4}{\beta^2} + 0.156\frac{1}{\beta^3}& \beta>2,
\end{cases}
\end{equation}
where $y=\operatorname{coth}\,\beta-1/\beta$.
We compare this analytical result and simulated values of $\chi_m$ with the Monte Carlo simulation in Fig.~\ref{fig:bergluscher}.
\begin{figure}[h!]
    \centering
      % \includegraphics[width=0.45\textwidth]{int_energy-0.png}
      % \includegraphics[width=0.45\textwidth]{int_energy-1.png}
      \includegraphics[width=0.8\textwidth]{berg_luscher.png}
      \caption{\label{fig:bergluscher} Comparison with \cite{berg1981}. First panel: internal energy compared with analytic  energy (Eq.~\ref{eq:analytic_energy}). Second panel: magnetic susceptibility compared with literature values.}
\end{figure}
The slight discrepancy between numerical results in the cross-over region provides an estimate of unquantified systematic uncertainties.

@@

\section{Summary}

The topological behavior of the non-linear sigma model has been a topic of debate for several decades. Berg and L\"uscher \cite{berg1981} originally highlighted the discrepancy between the renormalization group hypothesis, that the topological susceptibility is zero in the continuum limit, and the numerical results of nonperturbative calculations, supported by a semi-classical instanton picture, which indicate that the topological susceptibility is ill-defined in the continuum limit. They suggested three possible causes:
\begin{enumerate}[itemsep=-0.5em]
    \item The definition of the topological charge does not scale to the continuum.
    \item There are ultraviolet divergences.
    \item There is no reasonable continuum limit.
\end{enumerate}
Our results at finite gradient flow time, for both $\theta = 0$ and at non-zero $\theta$,  support the numerical results of \cite{bietenholz2018} at $\theta=0$ and indicate the presence of a divergence in the continuum limit. Pre-reflectively, this may suggest that ultraviolet divergences are not the origin of this divergence, because the flow suppresses ultraviolet fluctuations and introduces no new divergences in the renormalized NLSM \cite{makino2015a}. The story of this divergence is, however, more subtle and is consistent with divergences at finite flow time. The topological susceptibility (and all other even moments of the topological charge) are divergent in the continuum limit, differences between cumulants of the topological charge are finite \cite{Nogradi:2012dj}. This indicates that there is a single, non-perturbatively generated divergence that can be removed by introducing a single counter term in the Lagrangian, proportional to the identity. In the absence of any such counter-term, the NLSM at zero flow time is not completely renormalized, and introducing the flow cannot remove this divergence. One future application of the gradient flow to the study of the topology of the NLSM, could be, however, to define a nonperturbative renormalization scheme for this counter term, and study the corresponding running of its coupling.

\section{Acknowledgments}

We thank Daniel Nogradi for enlightening discussions on the NLSM at nonzero $\theta$ during this conference, and for drawing our attention to  \cite{Nogradi:2012dj}.
